### 2013-07-06 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding overlap data to document

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA Verification\ }
\label{sec:parallel_verification}
Before we consider the parallel performance of the MCSA algorithm, we
must first verify that the algorithm generates the correct solution in
parallel with respect to reference solutions provided by production
Krylov methods. To verify correctness, we use solutions the neutron
diffusion system presented in
Appendix~\ref{chap:diffusion_problem}. As the reference computation,
serial results will be used from the conjugate gradient solver in
Belos, a linear solvers package in the Trilinos scientific computing
libraries \citep{heroux_overview_2005}. In identical fashion to the
verification for the $SP_N$ equations performed in
\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
diffusion problem in parallel using multiple cores on a desktop
machine. For each parallel solution, the solution at each grid point
in the system will be compared to the solution at the same grid point
in the serial reference case with the minimum and maxiumum absolute
differences of these values reported.

@@

%%---------------------------------------------------------------------------%%
\section{Leadership-Class Parallel Scaling Studies\ }
\label{sec:leadership_scaling_studies}
Using the parallel performance metrics presented in the previous
section, we will next performance various scaling studies using an
implementation of the parallel MCSA algorithm to assess the quality of
the algorithm using the neutron diffusion problem presented in
Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
large problems at high levels of concurrency will be solved at a
leadership-class computing facility. In this section, we will outline
the parallel machine on which the calculations were performed and then
present a series of numerical studies that identify both strengths and
weaknesses of the parallel algorithm. Both pure domain decomposition
and MSOD parallel performance will be explored and compared to
performance results for the same problems using a set of production
parallel Krylov solvers. For the scaling results presented here, all
parallel Monte Carlo results were computed using the Monte Carlo
Linear Solvers\footnote{MCLS is an open source library and can be
  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
library generated specifically for this work while the conjugate
gradient and GMRES Krylov solvers used are part of the Belos linear
solver package within the Trilinos framework
\citep{heroux_overview_2005}. In general, the Belos package produces
solutions to the neutron diffusion problem over an order of magnitude
faster than the MCLS library for serial computations. Therefore, when
we directly compare these two codes, the MCLS parallel efficiencies
will appear artificially inflated relative to the Belos efficiencies
due to the fact that the MCLS ratio of work to communication will be
much larger than that for the Belos calculations.

@@

For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will isolate
specifically the performance of Brunner and Brantley's algorithm as
applied to the Neumann-Ulam algorithm. For the strong scaling study,
the global problem size was fixed at \sn{1.6}{7} DOFs for all solvers
and the 16 core run used as the base case. MCSA was used again with
one stochastic history used per DOF in the Neumann-Ulam solve to
compute the correction for the
acceleration. Figure~\ref{fig:titan_pure_strong} gives the results of
the this study with the given absolute efficiencies computed using
Eq~(\ref{eq:strong_scaling_absolute}).

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition strong scaling absolute
      efficiency.}  \textit{Super-linear speed-up is from memory
      thrashing in the base case. MCLS is an order of magnitude slower
      arithmetically, causing the higher efficiency. GMRES executes
      more operations than conjugate gradient, also creating a higher
      efficiency.}}
  \label{fig:titan_pure_strong}
\end{figure}

There are several key features of Figure~\ref{fig:titan_pure_strong}
worth noting. First, up to around 5,000 cores, all methods experience
a super-linear improvement in parallel efficiencies. For all methods,
this increase is not a function of the algorithmic efficiency as those
computed using Eq~(\ref{eq:algorithmic_efficiency}) remained at unity,
meaning that all computations converged in the same number of
iterations relative to the base case\footnote{When the algorithmic
  efficiency remains at unity for all calculations, the implementation
  efficiency is equivalent to the absolute efficiency and therefore is
  not presented here.}. Instead, what is happening here and was
observed by Brunner and Brantley in their work is that the large
problem used to scale to O(10,000) cores was large enough that memory
thrashing was occurring in the 16 core base case. As the number of
cores in the strong scaling study was increased and the local problem
size decreased, more of the local problem fit into the cache of the
system, thus causing improved serial performance and therefore
enhanced runtimes. In this region of super-linear efficiencies, we
also observe larger efficiencies for GMRES as compared to conjugate
gradient. This occurs due to the fact that conjugate gradient has a
static set of operations performed at every iteration while GMRES
operates on a continually growing subspace. In addition, the
orthogonalization operations required to generate that subspace
consume much more time than the basic conjugate gradient operations,
causing a larger serial runtime and subsequently higher
efficiencies. For the same reasons, we also observe higher
efficiencies for the MCSA implementation due to the serial performance
of the implementation.

@@

Next, we consider the weak scaling performance of the algorithm. In
this case, we fix the local problem size at \sn{4}{4} DOFs and
increase the number of processors used in the
simulation. Figure~\ref{fig:titan_weak_absolute} gives the weak
scaling absolute efficiencies for this problem as computed by
Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
when comparing MCSA to the Krylov methods are the same with MCSA and
GMRES scaling better than the conjugate gradient method due to the
increased amount of arithmetic work on-process. In addition, it is
important to note that the weak scaling of MCSA when properly
parallelized remains competitive with the Krylov methods keeping in
mind the artificial inflation of efficiencies.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition weak scaling absolute
      efficiency.} \textit{MCLS is an order of magnitude slower
      arithmetically, causing the higher efficiency. GMRES executes
      more operations than conjugate gradient, also creating a higher
      efficiency.}}
  \label{fig:titan_weak_absolute}
\end{figure}

As an additional means of comparison between MCSA and the Krylov
methods, we compute the algorithmic efficiencies of the calculations
using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
the Krylov algorithm improves as a function of the number of cores in
the problem due to the fact that the number of iterations required to
converge decreased from 20 in the base case to 19 and then to 18 for
the final set of computations. When this reduction in iterations is
used to account for the observed runtimes and computed absolute
efficiencies for weak scaling we end up with the implementation
efficiencies computed using Eq~(\ref{eq:implementation_efficiency})
given in Figure~\ref{fig:titan_weak_implementation}. Because MCSA
maintains the same iterative performance as a function of global
problem size in the weak scaling study, the implementation
efficiencies compare more favorably to the Krylov methods. In effect,
we are measuring the parallel performance of a single iteration for
each method by looking at these implementation efficiencies. As the
global problem size increases, MCSA is less prone to a drop in
parallel efficiency. When optimization has been completed and runtimes
potentially approach those of the Krylov methods, this comparison may
not be as favorable. Overall, weak scaling performance the parallel
MCSA algorithm is excellent for the pure domain decomposition case
with absolute parallel efficiencies of over 80\% observed at over
32,000 cores.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
  \end{center}
  \caption{\textbf{Binary tree vs. master/slave communication scheme
      absolute efficiency.} \textit{Pure domain decomposition absolute
      strong scaling. The master/slave scheme will not scale beyond
      $O(1,000)$ cores due to load imbalance.}}
  \label{fig:titan_strong_bvsm}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
  \end{center}
  \caption{\textbf{Binary tree vs. master/slave communication scheme
      absolute efficiency.}  \textit{Pure domain decomposition
      absolute weak scaling. The master/slave scheme will not scale
      beyond $O(1,000)$ cores due to load imbalance.}}
  \label{fig:titan_weak_bvsm}
\end{figure}

@@

@@

\subsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsec:msod_decomposition}

>>>>

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA Verification\ }
\label{sec:parallel_verification}
Before we consider the parallel performance of the MCSA algorithm, we
must first verify that the algorithm generates the correct solution in
parallel with respect to reference solutions provided by production
Krylov methods. To verify correctness, we use solutions to the neutron
diffusion system presented in
Appendix~\ref{chap:diffusion_problem}. As the reference computation,
serial results will be used from the conjugate gradient solver in
Belos, a linear solvers package in the Trilinos scientific computing
libraries \citep{heroux_overview_2005}. In identical fashion to the
verification for the $SP_N$ equations performed in
\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
diffusion problem in parallel using multiple cores on a desktop
machine. For each parallel solution, the solution at each grid point
in the system will be compared to the solution at the same grid point
in the serial reference case with the minimum and maxiumum absolute
differences of these values reported.

@@

%%---------------------------------------------------------------------------%%
\section{Leadership-Class Parallel Scaling Studies\ }
\label{sec:leadership_scaling_studies}
Using the parallel performance metrics presented in the previous
section, we will next performance various scaling studies using an
implementation of the parallel MCSA algorithm to assess the quality of
the algorithm using the neutron diffusion problem presented in
Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
large problems at high levels of concurrency will be solved at a
leadership-class computing facility. In this section, we will outline
the parallel machine on which the calculations were performed and then
present a series of numerical studies that identify both strengths and
weaknesses of the parallel algorithm. Both pure domain decomposition
and MSOD parallel performance will be explored and compared to
performance results for the same problems using a set of production
parallel Krylov solvers. For the scaling results presented here, all
parallel Monte Carlo results were computed using the Monte Carlo
Linear Solvers\footnote{MCLS is an open source library and can be
  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
library generated specifically for this work while the conjugate
gradient and GMRES Krylov solvers used are part of the Belos linear
solver package within the Trilinos framework
\citep{heroux_overview_2005}. In general, the Belos package produces
solutions to the neutron diffusion problem nearly two orders of
magnitude faster than the MCLS library for serial computations using
the conjugate gradient solver. Therefore, when we directly compare
these two codes, the MCLS parallel efficiencies will appear
artificially inflated relative to the Belos efficiencies due to the
fact that the MCLS ratio of work to communication will be much larger
than that for the Belos calculations. For every scaling study
presented, the calculations were performed 3 times with the average
values reported for wall time and used for the efficiency
computations.

@@

For the first set of scaling studies, we consider the parallel MCSA
algorithm where MSOD has not been used in the parallel Neumann-Ulam
algorithm. We refer to this case as pure domain decomposition where no
overlap or replication has been leveraged. Such a test will isolate
specifically the performance of Brunner and Brantley's algorithm as
applied to the Neumann-Ulam algorithm. For the strong scaling study,
the global problem size was fixed at \sn{1.6}{7} DOFs for all solvers
and the 16 core run used as the base case. MCSA was used again with
one stochastic history used per DOF in the Neumann-Ulam solve to
compute the correction for the acceleration with the remainder of the
problems and solver parameters the same as those for the verification
calculation given in Table~\ref{tab:parallel_verification_parameters}
and Table~\ref{tab:parallel_mcsa_parameters}. As with all of the
scaling studies, 3 calculations for all data points were performed
with the average wall time reported and used for the efficiency
calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
time per iteration for these calculations while
Figure~\ref{fig:titan_pure_strong} gives the results of the this study
with the given absolute efficiencies computed using
Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
iteration due to the fact that MCSA converged on the solution in 22
iterations while the Krylov solvers converged in 19 iterations.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong_time.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition strong scaling wall time
      per iteration.} \textit{Wall time is reported per iteration
      because the methods converged in different numbers of
      iterations. MCLS is over an order of magnitude slower
      arithmetically than the Krylov solvers. GMRES executes more
      operations than conjugate gradient and is therefore slower as
      well.}}
  \label{fig:titan_pure_strong_time}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition strong scaling absolute
      efficiency.}  \textit{Super-linear speed-up is from memory
      thrashing in the base case. MCLS is over an order of magnitude
      slower arithmetically, causing the higher efficiency. GMRES
      executes more operations than conjugate gradient, also creating
      a higher efficiency.}}
  \label{fig:titan_pure_strong}
\end{figure}

There are several key features of
Figure~\ref{fig:titan_pure_strong_time} and
Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
5,000 cores, all methods experience a super-linear improvement in
parallel efficiencies. For all methods, this increase is not a
function of the algorithmic efficiency as those computed using
Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
all computations converged in the same number of iterations relative
to the base case\footnote{When the algorithmic efficiency remains at
  unity for all calculations, the implementation efficiency is
  equivalent to the absolute efficiency and therefore is not presented
  here.}. Instead, what is happening here and was observed by Brunner
and Brantley in their work is that the large problem used to scale to
O(10,000) cores was large enough that memory thrashing was occurring
in the 16 core base case. As the number of cores in the strong scaling
study was increased and the local problem size decreased, more of the
local problem fit into the cache of the system, thus causing improved
serial performance and therefore enhanced runtimes. In this region of
super-linear efficiencies, we also observe larger efficiencies for
GMRES as compared to conjugate gradient. This occurs due to the fact
that conjugate gradient has a static set of operations performed at
every iteration while GMRES operates on a continually growing
subspace. In addition, the orthogonalization operations required to
generate that subspace consume much more time than the basic conjugate
gradient operations, causing a larger serial runtime and subsequently
higher efficiencies. For the same reasons, we also observe higher
efficiencies for the MCSA implementation due to the serial performance
of the implementation.

@@

Next, we consider the weak scaling performance of the algorithm. In
this case, we fix the local problem size at \sn{4}{4} DOFs and
increase the number of processors used in the
simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall
time per iteration and Figure~\ref{fig:titan_weak_absolute} gives the
weak scaling absolute efficiencies for this problem as computed by
Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
when comparing MCSA to the Krylov methods are the same with MCSA and
GMRES scaling better than the conjugate gradient method due to the
increased amount of arithmetic work on-process. In addition, it is
important to note that the weak scaling of MCSA when properly
parallelized remains competitive with the Krylov methods keeping in
mind the artificial inflation of efficiencies.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_weak_time.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition weak scaling wall time
      per iteration.} \textit{Wall time is reported per iteration
      because the methods converged in different numbers of
      iterations. MCLS is over an order of magnitude slower
      arithmetically than the Krylov solvers. GMRES executes more
      operations than conjugate gradient and is therefore slower as
      well.}}
  \label{fig:titan_pure_weak_time}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
  \end{center}
  \caption{\textbf{Pure domain decomposition weak scaling absolute
      efficiency.} \textit{MCLS is an over order of magnitude slower
      arithmetically, causing the higher efficiency. GMRES executes
      more operations than conjugate gradient, also creating a higher
      efficiency.}}
  \label{fig:titan_weak_absolute}
\end{figure}

As an additional means of comparison between MCSA and the Krylov
methods, we compute the algorithmic efficiencies of the calculations
using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
the Krylov algorithm improves as a function of the number of cores in
the problem due to the fact that the number of iterations required to
converge decreased from 20 in the base case to 19 and then to 18 for
the final set of weak scaling computations. When this reduction in
iterations is used to account for the observed runtimes and computed
absolute efficiencies for weak scaling we end up with the
implementation efficiencies computed using
Eq~(\ref{eq:implementation_efficiency}) given in
Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
same iterative performance as a function of global problem size in the
weak scaling study, the implementation efficiencies compare more
favorably to the Krylov methods. Using the implementation efficiency,
we are effectively measuring the parallel performance of a single
iteration for each method by looking at these implementation
efficiencies. As the global problem size increases, MCSA is less prone
to a drop in parallel efficiency. When optimization has been completed
and runtimes potentially approach those of the Krylov methods, this
comparison may not be as favorable. Overall, weak scaling performance
the parallel MCSA algorithm is excellent for the pure domain
decomposition case with absolute parallel efficiencies of over 80\%
observed at over 32,000 cores.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
  \end{center}
  \caption{\textbf{Binary tree vs. master/slave communication scheme
      strong scaling absolute efficiency.} \textit{The master/slave
      scheme will not scale beyond $O(1,000)$ cores due to load
      imbalance.}}
  \label{fig:titan_strong_bvsm}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
  \end{center}
  \caption{\textbf{Binary tree vs. master/slave communication scheme
      weak scaling absolute efficiency.}  \textit{The master/slave
      scheme will not scale beyond $O(1,000)$ cores due to load
      imbalance.}}
  \label{fig:titan_weak_bvsm}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap.pdf}
  \end{center}
  \caption{\textbf{Strong scaling absolute efficiency with varying
      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
      of overlap were used as the problem had a root-mean-squared
      number of states of 2.63 for each history.}}
  \label{fig:titan_strong_overlap}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap_diff.pdf}
  \end{center}
  \caption{\textbf{Strong scaling absolute efficiency difference with
      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
      and 5 levels of overlap are compared here with the absolute
      difference of the absolute efficiencies presented.}}
  \label{fig:titan_strong_overlap_diff}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap.pdf}
  \end{center}
  \caption{\textbf{Weak scaling absolute efficiency with varying
      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
      of overlap were used as the problem had a root-mean-squared
      number of states of 2.63 for each history.}}
  \label{fig:titan_weak_overlap}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap_diff.pdf}
  \end{center}
  \caption{\textbf{Weak scaling absolute efficiency difference with
      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
      and 5 levels of overlap are compared here with the absolute
      difference of the absolute efficiencies presented.}}
  \label{fig:titan_weak_overlap_diff}
\end{figure}

@@
